---
title: "Oil Prices Visualizations"
author: "Kamau Kamau"
date: "29/03/2021"
output:
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F,warning = F,message =F)
```

# High prices of oil in Kenya and the role of EPRA

The kenyan prices has been increasing for the better part of 2021 and this is being felt across all Kenyan sectors. Oil prices theoretically have had a link to all economic sectors which is being translated into high costs of living for the Kenyan households. The main question on Kenyans Minds is why are these prices high and what role is the E playing in regulating them. Well, for one, the ERC is a government parastatal tasked with managing the prices of oil and its product here in Kenya so as t stabilize the economy. 

```{r,fig.width =10}
library(tidyverse)
library(readxl)
library(lubridate)
library(reshape2)
setwd("C:\\Users\\Maria\\Documents\\AESOPS\\Oil_price")
#dir()
df <- read_excel("Oil_prices_master_file.xlsx" )
df$Date <- as.Date(df$Date,format = "%d/%m/%Y")
df$Average <- (df$PMS+df$AGO+df$Kero)/3
df$ppb_ken <- (df$`Price Per Barell` * df$Exrates)/158.99
head(df) %>% knitr::kable()


```

```{r,fig.width =10,width.height =11}
library(ggthemes)
test <- as.Date("10/01/2021",format = "%d/%m/%Y")
data <- df  %>%  select(Date,ppb_ken,Average) 
names(data)<- c("Date","Price_int","Price_local")
data%>% melt(id.vars = "Date")%>% ggplot(aes(x = Date,y = value,color=variable))+geom_line()+
  theme(legend.position = "bottom")+scale_y_continuous(breaks = seq(0,120,10))+
     # adjust the x axis breaks
     scale_x_date(date_breaks = "1 years", date_labels = "%Y")+labs(title = "Movement of oil prices in Kenya vs international market")+ annotate(geom="text", x=test, y=50, label="International",
              color="red")+annotate(geom="text", x=test, y=110, label="Local",
              color="red")+theme_classic()+theme(legend.position = "bottom",axis.title = element_text(size = 14),
    axis.text = element_text(size = 12, face = "bold"), plot.title = element_text(hjust = 0.4))
```

# percentage change

```{r,fig.width =10,fig.height =6}
per_avg <- c()
per_bar <- c()

for (i in 1:124) {
  
  col = df$Average
  col1 = df$ppb_ken
  
  n = (col[i] -col[i+1])/col[i]
  per_avg <- c(per_avg,n)
  
  n1 = (col1[i] -col1[i+1])/col1[i]
  per_bar <- c(per_bar,n1)
  

  
}
test <- as.Date("10/01/2015",format = "%d/%m/%Y")
new <- data.frame(Average = per_avg,
                  Barell  =  per_bar)
new$Date <-df$Date
names(new) <-  c("Kenyan","International","Date")
new %>% melt(id.vars = "Date") %>% ggplot(aes(x = Date,color = variable,y = value))+geom_line()+
  theme(legend.position = "bottom")+labs(title = "Oil fluctuations")+
  theme(legend.position = "bottom")+scale_y_continuous(breaks = seq(0,120,1))+
     # adjust the x axis breaks
     scale_x_date(date_breaks = "1 years", date_labels = "%Y")+labs(title = "Percentage price changes")+theme_classic()+theme(legend.position = "bottom",axis.title = element_text(size = 14),
    axis.text = element_text(size = 12, face = "bold"), plot.title = element_text(hjust = 0.4))+annotate(geom="text", x=test, y=0.3, label="The kenyan prices are more stable",
              color="green")

```

